An Animated Visualization Method for Dynamic Scale-Free Networks\thanks{Supported by FAPESP, project 2017/05838-3}

### 2019-06-13 doc/paper.tex
todos os arquivos usados

\begin{document}
%
\title{Versinus: an Animated Visualization Method for Dynamic Scale-Free Networks\thanks{Supported by FAPESP, project 2017/05838-3}}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Renato Fabbri\orcidID{0000-0002-9699-629X} \and
Maria Cristina Ferreira de Oliveira\orcidID{0000-0002-4729-5104}}
%
\authorrunning{R. Fabbri and M. C. F. de Oliveira}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{University of São Paulo, São Carlos SP, Brazil\\
\email{renato.fabbri@gmail.com},
\email{cristina@icmc.usp.br}}
%\url{http://conteudo.icmc.usp.br/pessoas/cristina/}}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
% The abstract should briefly summarize the contents of the paper in
% 150--250 words.
Most real-world networks are dynamic, i.e., they change over time as nodes and links are added or removed, and their representation may also involve variations in the associated data.
There is a growing interest in modeling dynamic (or time-evolving, or longitudinal) networks and
a number of procedures have been proposed to visualize them.
  This article presents a visualization method for dynamic scale-free networks.
  Compliant methods were not found in other tools by the authors, which motivated this description and the development of a dedicated interface.
  Named Versinus, the visualization is specially useful in the observation of (in)stability of basic topological characteristics, and consists essentially in placing the most connected nodes (hubs and intermediary) along a sine curve, and the peripheral nodes along a separate segment, usually a line.
  Thus its name, from  Latin \emph{versus} (line) and \emph{sinus} (sine).
  The Versinus visualization can provide insight into network properties and is useful for observing its behavior, as described in this paper, after a formal presentation of the method and its associated software implementations.
\keywords{Network visualization
\and Dynamic networks
\and Longitudinal networks
\and Time-evolving networks
\and Evolving networks
\and Animated visualization
\and Complex networks
\and Data visualization.}
\end{abstract}
%
%
%
\section{Introduction}
The visualization of dynamic networks poses challenges related to data size and complexity which have merited various contributions~\cite{evo1,hajij,evo3,evo2}.
In fact, the meaningful visualization of dynamic networks is a well-known problem in Information Visualization and dynamic network analysis.
In this work, we present an approach for dynamic network visualization that consists in distributing the nodes along a fixed layout arranged in three segments associated with the three sectors (or partitions) of a scale-free network, namely for the hub, intermediary and peripheral nodes~\cite{stab}. The network dynamic features are expressed by means of changes in visual properties of the glyphs used to depict nodes and links, e.g., glyph height, width, color and shape, and line (arrow) disposition, thickness and color. Moreover, complementary global and local information is rendered in order to enable an acoustic depiction of features of interest.
The method has been called Versinus since its early usages in 2013, because the layout stabilized, after attempts with distinct arrangements, in a sine period and a line, which in Latin is \emph{sinus} (sine) and \emph{versus} (line). 
There are four software implementations of Versinus, two of them written in Python scripts and packages that yield animations through stop-motion. The other two are written in JavaScript, one focused on exploitation of static networks, the other is an experimental Visual Analytics interface, which extends the visual metaphor to the 3D space and provides additional analysis techniques such as principal component analysis (PCA) and resources for structure navigation\footnote{An instance of the interface is available at: http://rfabbri.vicg.icmc.usp.br:3000/evolution}.
The writing of this document was motivated by
the fact that no method was found by the authors that aims at visualizing dynamic scale-free networks through node-link diagrams.
Furthermore, the visualization method was employed in scientific research~\cite{thesis,stab}, and thus a careful mathematical and technological description of the procedures involved in applying Versinus to real networks is appropriate.

The article is organized as follows. The next subsections address related work and issues of nomenclature. Section~\ref{sec:des} describes the method in detail and sufficient formalism. Section~\ref{sec:imp} outlines the software implementations of Versinus and depicts the resulting animations. Section~\ref{sec:ins} is dedicated to revealing further insights obtained through the use of the method. Finally, Section~\ref{sec:con} assembles concluding commentaries and further work.

\subsection{Related work}
% on animated visualization
% on vis of evo nets
Although animation for the purposes of data analysis is often criticized in the data visualization literature, advantages have also been reported~\cite{anim,munzner,cog,ware}, e.g.: yielding visualizations more eye catching and taking advantage of the fact that humans are used to interacting with moving objects.
Dynamic network visualization has been tackled both through animated and static (often timeline-based) visualizations~\cite{hajij,brain,visAn,ego}, focusing on various global and local network features.
The contribution reported in this document is singular in that it consists basically of a layout which is fit for networks with scale-free characteristics, in association with some visualization cues which are implementation dependent. It results from extensive attempts to reach a satisfactory visualization by means theoretical and intuitive strategies and methodical verification of the result by appreciation of the entailed animations.
The procedures and conceptualizations involved in achieving the final visualization were also considered worth documenting, as they were not found elsewhere in the literature, and by virtue of the simple and genuine design. Its scientific relevance on a practical application has been demonstrated before~\cite{stab}.

\subsection{Remarks on nomenclature}
Dynamic networks are also called e.g. longitudinal networks, time-evolving networks, evolving networks, and time-varying networks~\cite{ev,tev,tv,lg}.
We chose to adopt the term ``dynamic network'' herein because it is simple, semantically meaningful and has been widely adopted.
Another major issue in current vocabulary is in the phrase ``sector of a network'' (e.g. of the hubs). In most cases, the sectors are also partitions (i.e. they are non-overlapping sets whose union is the complete superset), but it depends on the criteria for obtaining the sectors~\cite{stab}, thus the use of the more general term \emph{sector}.
Moreover, network science is highly multidisciplinary and context-dependent, and flourished from the ill-defined area of complex systems, favoring polysemy and synonymy. We strove to avoid such issues and keep a consistent usage of nomenclature.
%because they are usually not convenient in scientific literature.

\section{A description of the method}\label{sec:des}
Versinus generates a layout through placing the nodes in segments
with respect to their overall topological features and then representing
instantaneous topological features though glyph and arrow attributes.
The visualization may be thus split in three stages, described in the following subsections.

\subsection{Node placement}\label{sec:pla}
First, a network is constructed with all the activity of interest, i.e. the transactions such as all the email messages considered, or all the links in a data file.
The nodes are then ordered either by degree or strength (or in- or out-degree or strength), from greatest to smallest values.
Also from such global network, the hubs, intermediary and peripheral sectors are obtained. Two methods have been proven useful for obtaining sectors: 1) assuming fixed fractions of hubs (e.g. 5\%), of intermediary (e.g. 15\%) and peripheral nodes (e.g. 80\%); 2) by comparison of the network against the Erdös-Rényi model, a non-trivial procedure, with relevant analytical results, detailed elsewhere~\cite{stab}.

With such structures at hand, the nodes are placed along three segments, as depicted in Figure~\ref{fig:glob}.
The hubs are placed along the first half of a sine period.
The intermediary nodes are placed along the second half of the sine period.
The peripheral nodes are placed along an independent line.
The placement follows 
the basic parameters:
\begin{itemize}
  \item The amplitude $\alpha$ of the sine oscillation.
  \item The displacement $\Delta$ of the sine in the vertical axis.
  \item The horizontal margin $\mu$, i.e. the distance from the left border to the first hubs, assumed the same as the distance from the right border to last intermediary node.
  \item The positions $(x_0,y_0)$ and $(x_1,y_1)$, i.e. the endpoints of the segment for the peripheral sector.
\end{itemize}

\begin{figure}[!h]\centering
\includegraphics[width=.7\textwidth]{nodePositioning}
  \caption{The underlying layout of Versinus. Highly connected nodes are positioned from left to right on the sine segment (for the hubs and intermediary sectors), and from right to left on the upper line (for the periphery sector). The parameters are: amplitude $\alpha$ of the oscillation; vertical displacement $\Delta$ of the oscillation; horizontal margin $\mu$; and line endpoints $(x_0, y_0)$ and $(x_1, y_1)$.
   Further information is given in Section~\ref{sec:pla}.
  }\label{fig:glob}
\end{figure}

\noindent Some crafty remarks: fine-tuning the size (number of nodes) of each sector may come handy due to the node density and activity in each segment. Also, the node placed at the middle of the sine period may be a hub or an intermediary node, and this choice often impacts the clarity of the final visualization more than anticipated by intuition.
Finally, avoiding clutter with the hubs motivated both 
the positioning of the peripheral sector in the upper-right corner and
the placement of the most active peripheral nodes next to the intermediary sector, i.e. the ordering along the peripheral line.

\subsection{Drawing nodes, links and auxiliary information}
Additional concepts and quantities need to be defined for visualizing the evolving network as a data-driven animation.
A \emph{snapshot} $S$ consists of a number $\sigma$ time-contiguous transactions, e.g. 100 email messages, starting at an arbitrary message.
The separation $\eta$ between each snapshot is the number of transactions skipped from one snapshot to the next.
In summary, the evolving network is visualized through a sliding window of $\sigma$ transactions, updated at every $\eta$ transactions. To achieve the final animation, a transaction rate $\omega$ must also be defined.
Thus, the following parameters are specified:
\begin{itemize}
  \item The window $\sigma$ captured at each snapshot, expressed in transactions.
  \item The step size $\eta$ between each snapshot, expressed in transactions.
  \item The transaction rate $\omega$, expressed in transactions per second.
\end{itemize}

\noindent The sequence of snapshots may be expressed as $\{S_i^{\sigma, \eta}\}_0^{L}$, where $L$ is the smallest integer such that $L\times \eta + \sigma > T$ and $T$ is the total number of transactions. Notice that the final snapshot $S_L$ may have less than $\sigma$ transactions. Also, the rate $\omega$ is only applied when rendering the animation, e.g. if each snapshot $S_i$ yields an image, the stop motion animation will use $\omega/\eta$ images per second to achieve the desired rate of $\omega$ transactions per second.

@@

\noindent Finally, complementary information may be displayed as a persistent legend that reveals essential information such as $\sigma$, $\eta$, $\omega$ and $T$, and keeps track of the current snapshot, e.g. its number of nodes and edges.
Also, node-specific information may be displayed by blinking corresponding cues over the nodes periodically, displaying them with a vertical displacement from the node, or by writing them to a specific widget when requested by the analyst on an interactive interface, i.e. on-demand.
Figure~\ref{fig:final} exemplifies two of the early flavors of Versinus settings.

\begin{figure}[!h]\centering
    \subfloat[]{{\includegraphics[width=.46\textwidth, height=4.3cm]{versinus_} }}%
    \quad
  \subfloat[]{{\includegraphics[width=.46\textwidth, height=4.3cm]{im31} }}%
    \caption{Two frames of dynamic network animations achieved using the Versinus method.
    Both are achieved through dedicated Python scripts and package modules.
    In (a) the individual degrees in the snapshot are blinking over the nodes, and the nodes are counted in auxiliary lines along the bottom and paralel to the periphery line. Current image/snapshot is 911 and has 116 nodes and 262 links.
   In (b), a cleaner version is employed to display a snapshot in which the only hub is the overall hub, the second overall hub is a peripheral node, and most nodes have high clustering coefficient. The legend indicates that the total number of transactions is $T=3,998$, the window size $\sigma=40$, the step size is $\eta=40$, the transaction rate is $\omega=4$, the current snapshot starts at transaction 1,240 and has 13 nodes and 15 links.
   The low-quality images are typical of the early implementations of Versinus, developed to assist the investigations of the researchers directly related to its development.
   Further information is given in Section~\ref{sec:des}.
   }%
    \label{fig:final}%
\end{figure}

\subsection{Observation of the evolving network}
Within the settings described, one may notice a number of network characteristics: do nodes (or a specific node) change their sectors often? Is the overall intermediary sector in greater activity with the hubs or periphery?
Does higher clustering really occur most often in the less connected nodes?
Are there very unbalanced nodes with respect to in and out degree (or strength)?
Is the number of nodes and edges stable along the network evolution?
In fact, Versinus was partly designed to inspect network attributes due to the remarkable stability of their overall characteristics reported in~\cite{stab}.
In this context, it is natural to ask if such stability is also noticeable in individual nodes or in connectivity patterns. Although the overall characteristics may be regarded as surprisingly stable, the stability of network components is very network-dependent. E.g.,  as suggested by~\cite{barabasiEvo}, all hubs were found to have intermittent activity in social networks, except for the smallest networks analyzed.

\section{Software implementations}\label{sec:imp}
Currently, there are four software implementations of Versinus known to the authors.
For organization and brevity, these are all linked in\footnote{Pivot repository for Versinus resources such as implementations, public instances of interactive interfaces, and videos: https://github.com/ttm/versinus}. 
Two of them are written in Python: one is a script that renders animations as in the Figure~\ref{fig:final}
(a), with more cues of local structure. The other is a library that renders animations as in the Figure~\ref{fig:final} (b), simpler and with more cues of the overall structure, and is capable of rendering sonifications of the dynamic networks, yielding ``audiovisualizations'', i.e. synced sonic and visual mappings of the dynamic networks.
The other two implementations are in JavaScript and use WebGL. One is, by default, for static (non-dynamic) networks, implemented as part of the ccNetViz library, that aims at being as computationally inexpensive as possible~\cite{cell}.
The other is an experimental 3D Visual Analytics application in which Versinus is combined with other analytical techniques,
such as principal component analysis, overview first - focus on demand, and simultaneous highlighting of corresponding structures across images as required by users.
These last two implementations are illustrated in Figure~\ref{fig:extra}.

\begin{figure}[!h]\centering
    \subfloat[Versinus in ccNetViz]{{\includegraphics[width=.46\textwidth, height=2.5cm]{ccNetViz} }}%
    \qquad
  \subfloat[Experimental Visual Analytics]{{\includegraphics[width=.46\textwidth]{visAn} }}%
    \caption{Two implementations of Versinus in JavaScript using WebGL. (a) illustrates the implementation that is available within ccNetViz, a library purposed to being computationally inexpensive. This implementation handles static networks by default. (b) is an image achieved using an experimental Visual Analytics application.
    It renders the animations in real time, with multiple controls for inspecting the network structure and controlling the animation, in association with other analysis techniques. The sine oscillates both in the vertical and depth axes, as does the periphery segment. Further information is given in Section~\ref{sec:imp}.}
    \label{fig:extra}
\end{figure}

\subsection{Network sonification}\label{sson}
The implementation illustrated in Figure~\ref{fig:final} (b) includes a sonification routine for the evolving network.
Results are of limited effectiveness in the sense that the sonification does not enable users to grasp network features with ease,
specially if they are not musically trained.
Nevertheless, it made the (audio)visualization more engaging, as it yields exotic media,
and it is the first time the authors appreciated a sonification of an evolving network (maybe even of any network),
and is thus probably a unique multimedia representation of networks.
Accordingly, we understand suitable to provide a short description of the sonification as it was left implemented,
i.e. in its final version,
even though a full understanding of the summary requires knowledge of musical theory.
Four sonic/musical voices are used, two of them are noises and the other two have defined pitch. The lower noise has central frequency proportional to first hub connectivity,
which is the node positioned at the extreme left on the layout, on the starting point of the sine, and its frequency is updated every two beats.
The higher noise has central frequency proportional to the second hub, the node next to the first hub, and its frequency is updated at each beat.
The third hub connectivity is proportional to the lower note,
which is incident once a beat, and has its note value updated at every beat.
The fourth hub connectivity is proportional to the higher note, which is incident twice every half-beat, is updated at every note, and is only heard on the second half-beat of every tempo.
Figure~\ref{fson} illustrates such disposition of sonic lines (or musical voices).

\begin{figure}[!h]\centering
\includegraphics[width=.9\textwidth]{sonif}
  \caption{Visual representation of the sonification implemented for the dynamic networks.
  $c_f$ and $f_0$ stands for \emph{center frequency} and \emph{fundamental frequency}, respectively.
  Further details are given in Section~\ref{sson}.
  }\label{fson}
\end{figure}

\subsection{Animations produced and available}\label{sec:vid}
A number of animations were obtained through the routines described.
These encompassed different networks, settings ($\alpha$, $\sigma$, $\eta$,
etc), and different flavors of Versinus.
Some of these visualizations are available in six playlists of Youtube videos.
For organization and brevity, they are available at~\footnote{Pivot repository for Versinus resources such as implementations, public instances of interactive interfaces, and videos: https://github.com/ttm/versinus} together with other Versinus-related resources.

\subsection{An experimental Visual Analytics interface}
This section exemplifies the employment of the Versinus method
within a more elaborate framework.
Let us consider an interactive visual interface, with which the user may parametrize Versinus and then navigate through the entailed dynamic network.

\begin{figure}[!h]\centering
\includegraphics[width=\textwidth]{esfNetVis___}
  \caption{
  }\label{esfnetvis}
\end{figure}

\begin{figure}[!h]\centering
\includegraphics[width=\textwidth]{esfNetVis2___}
  \caption{
  }\label{esfnetvis}
\end{figure}

\begin{figure}[!h]\centering
\includegraphics[width=\textwidth]{esfNetVis3}
  \caption{
  }\label{esfnetvis}
\end{figure}

\begin{figure}[!h]\centering
\includegraphics[width=\textwidth]{esfNetVis4}
  \caption{
  }\label{esfnetvis}
\end{figure}
One basic feature of the method is the constant representation of
nodes in each of the hubs, intermediary and peripheral sectors.
Nodes are positioned according to such classification regarding all
the activity of interest, and in each snapshot visual cues map the
classification within the snapshot.
Following the analytic methods used in~\cite{stab}, it may be crucial
to know how the relative sizes of the sectors vary along time, i.e. along
the sequence of snapshots.

\section{Insights gathered}\label{sec:ins}
This section describes insights gathered by the authors,
and should be further verified in subsequent work and using
suitable validation methods.
The legitimacy of the exposition herein given is based on the reasonable aspect of the assertions, and on the facts that the authors developed Versinus, used the method extensively, and the scientific contribution which it made possible~\cite{stab}.

>>>>

\begin{document}
%
\title{An Animated Visualization Method for Dynamic Scale-Free Networks\thanks{Supported by FAPESP, project 2017/05838-3}}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Renato Fabbri\orcidID{0000-0002-9699-629X} \and
Maria Cristina Ferreira de Oliveira\orcidID{0000-0002-4729-5104}}
%
\authorrunning{R. Fabbri and M. C. F. de Oliveira}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{University of São Paulo, São Carlos SP, Brazil\\
\email{renato.fabbri@gmail.com},
\email{cristina@icmc.usp.br}}
%\url{http://conteudo.icmc.usp.br/pessoas/cristina/}}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
% The abstract should briefly summarize the contents of the paper in
% 150--250 words.
Most real-world networks are dynamic, i.e., they change over time as nodes and links are added or removed, and their representation may also involve variations in the associated data.
There is a growing interest in modeling dynamic (or time-evolving, or longitudinal) networks and
a number of procedures have been proposed to enable their visualization.
  This article presents a visualization method for dynamic scale-free networks.
  Named Versinus, the method consists of a specialized combination of layout and further visual cues
  useful in the observation of basic topological characteristics.
  The layout consists essentially in placing the most connected nodes (hubs and intermediary)
  along a sine curve, and the peripheral nodes along a separate segment, usually a line,
  thus its name, from the Latin \emph{versus} (line) and \emph{sinus} (sine).
  The visual cues yield evident both local and global key-features.
  A compliant method was not found in the scientific literature or among existing tools by the authors.
  Therefore, after a formal presentation of Versinus and its associated software implementations,
  a simple case study is described in order to provide insights into the potential exploitation of the method
  in analysing dynamic scale-free networks.
  % which motivated this description and the development of a dedicated interface.
\keywords{Network visualization
\and Scale-free networks
\and Dynamic networks
\and Animated visualization
\and Complex networks
\and Data visualization.}
\end{abstract}

\section{Introduction}
The visualization of dynamic networks poses challenges related to data size and
complexity which have merited various
contributions~\cite{evo1,evo3,hajij,evo2}.  In fact, the meaningful
visualization of dynamic networks is a well-known problem in Information
Visualization and dynamic network analysis.
In this work, we present a peculiar
method for dynamic network visualization which consists in distributing the
nodes along a fixed layout and mapping the network evolution through changes in
glyphs corresponding to nodes and links.
The layout is arranged in three segments, each associated with one of the three
sectors (or partitions) of a scale-free network, namely the hub,
intermediary and peripheral nodes~\cite{stab}.
Local features of the network are mapped to the visual properties of the glyphs,
e.g., glyph height, width, color and shape, and line (arrow)
thickness and color. Moreover, complementary global and local
information is rendered in order to enable the observation of basic aspects
of the dynamic network.
The method has been called Versinus since its early usages in 2013,
because the layout settled, after attempts with distinct arrangements, in a
sine period and a line, which in Latin is \emph{sinus} (sine) and \emph{versus}
(line).
No compliant method was found by the authors
for visualizing dynamic scale-free networks through node-link
diagrams.  Furthermore, the visualization method was employed in scientific
research~\cite{thesis,stab} and has received a number of software implementations.

% There are four software implementations of Versinus, two of them
% written through Python scripts and packages, and yield animations through
% stop-motion. The other two are written in JavaScript, one is focused on the
% exploitation of static networks, the other is an experimental Visual Analytics
% interface, which extends the visual metaphor to the 3D space and provides
% additional analysis techniques such as principal component analysis (PCA) and
% resources for structure navigation\footnote{Central repository for Versinus resources such as implementations, public instances of interactive interfaces, and videos: https://github.com/ttm/versinus}.
% The writing of

The article is organized as follows. The next subsections address related work and nomenclature. Section~\ref{sec:des} describes the method. Section~\ref{sec:imp} outlines the software implementations of Versinus and depicts the resulting animations and interfaces. Section~\ref{visAn} describes a simple case study in order to provide insights about the potentialities of the method. Section~\ref{sec:ins} is dedicated to hypotheses provided through the use of the method. Finally, Section~\ref{sec:con} assembles concluding commentaries and further work.

\subsection{Related work}
% on animated visualization
% on vis of evo nets
Although animation for the purposes of data analysis is often criticized in the data visualization literature, advantages have also been reported~\cite{anim,munzner,cog,ware}, e.g.: for eye catching visualizations and potential advantages yield by the fact that humans are used to interacting with moving objects.
Dynamic network visualization has been tackled both through animated and static (often timeline-based) visualizations~\cite{hajij,brain,visAn,ego}, focusing on various global and local network features.
The contribution reported in this document is possibly singular in that it consists basically of a layout which was designed for networks with scale-free characteristics, in association with some visual cues.
It results from extensive attempts to reach a satisfactory visualization by means theoretical and intuitive strategies and methodical verification of the result by appreciation of the entailed animations.
% The procedures and conceptualizations involved in achieving the final visualization were also considered worth documenting, as they were not found elsewhere in the literature, and by virtue of the simple and genuine design. Its scientific relevance on a practical application has been demonstrated before~\cite{stab}.

\subsection{Remarks on nomenclature}
Dynamic networks are also called e.g. longitudinal networks, time-evolving networks, evolving networks, and time-varying networks~\cite{ev,tev,tv,lg}.
We adopted the term ``dynamic network'' herein because it is simple, semantically meaningful and has been widely adopted.
Another major issue in current vocabulary arises when considered the hub, intermediary and peripheral sectors.
In most cases, the sectors are also partitions (i.e. they are non-overlapping sets whose union is the complete superset), but it depends on the criteria for obtaining the sectors~\cite{stab}, thus the use of the more general term \emph{sector}.
Moreover, network science is highly multidisciplinary and context-dependent, and flourished from the vaguely-defined area of complex systems, favoring polysemy and synonymy. We strove to avoid such issues and to employ a consistent.
%because they are usually not convenient in scientific literature.

\section{A description of the method}\label{sec:des}
Versinus layout consists basically in placing the nodes in segments
with respect to their overall topological features and then representing
instantaneous topological features through glyph attributes.
Additional, implementation-dependent, information is often provided.
The achievement of the visualization may be thus split in three stages,
described in the following subsections.

\subsection{Node placement}\label{sec:pla}
First, a network is constructed with all the activity of interest, i.e. the transactions such as all the email messages considered, or all the links in a data file.
The nodes are then ordered either by degree or strength (or in- or out-degree or strength), from greatest to smallest values.
Also from such global network, the hubs, intermediary and peripheral sectors are obtained. Two methods have proven useful for obtaining the sectors: 1) assuming fixed fractions of hubs (e.g. 5\%), of intermediary (e.g. 15\%) and peripheral nodes (e.g. 80\%); 2) comparison of the network against the Erdös-Rényi model, a non-trivial procedure, with relevant analytical results, detailed in~\cite{stab}.

With such structures at hand, the nodes are placed along three segments, as depicted in Figure~\ref{fig:glob}.
The hubs are placed along the first half of a sine period.
The intermediary nodes are placed along the second half of the sine period.
The peripheral nodes are placed along an independent line.
These are the basic parameters:
\begin{itemize}
  \item The amplitude $\alpha$ of the sine oscillation.
  \item The displacement $\Delta$ of the sine in the vertical axis.
  \item The horizontal margin $\mu$, i.e. the distance from the left border to the first hub, assumed to be the same as the distance from the right border to last intermediary node.
  \item The positions $(x_0,y_0)$ and $(x_1,y_1)$, i.e. the endpoints of the segment for the peripheral sector.
\end{itemize}

\begin{figure}[!h]\centering
\includegraphics[width=.7\textwidth]{nodePositioning}
  \caption{The underlying layout of Versinus. Highly connected nodes are positioned from left to right on the sine segment (for the hubs and intermediary sectors), and from right to left on the upper line (for the periphery sector). The parameters are: amplitude $\alpha$ of the oscillation; vertical displacement $\Delta$ of the oscillation; horizontal margin $\mu$; and line endpoints $(x_0, y_0)$ and $(x_1, y_1)$.
   Further information is provided in Section~\ref{sec:pla}.
  }\label{fig:glob}
\end{figure}

\noindent Remarks: fine-tuning the size (number of nodes) of each sector may come handy due to the node density and activity in each segment. Also, the node placed at the middle of the sine period may be a hub or an intermediary node, and this choice often impacts the clarity of the final visualization more than anticipated by intuition, specially in small networks.
Finally, avoiding clutter with the hubs motivated both 
the positioning of the peripheral sector in the upper-right corner and
the placement of the most active peripheral nodes next to the intermediary sector, i.e. the ordering along the peripheral line.

\subsection{Drawing nodes, links and auxiliary information}
Let a \emph{snapshot} $S$ consist of a number $\sigma$ time-contiguous transactions, e.g. 100 email messages, starting at an arbitrary transaction.
The separation $\eta$ between each snapshot is the number of transactions skipped from one snapshot to the next.
The evolving network is thus conceptualized as a sliding window of $\sigma$ transactions, which progresses by skipping $\eta$ transactions at each step.
To achieve the final animation, a transaction rate $\omega$ must also be defined.
Thus, the following parameters are defined:
\begin{itemize}
  \item The window size of $\sigma$ captured at each snapshot, expressed in transactions.
  \item The step size $\eta$ between each snapshot, expressed in transactions.
  \item The transaction rate $\omega$, expressed in transactions per second.
\end{itemize}

\noindent The sequence of snapshots may be expressed as $\{S_i^{\sigma, \eta}\}_0^{L}$, where $L$ is the smallest integer such that $L\times \eta + \sigma > T$ and $T$ is the total number of transactions. Notice that the final snapshot $S_L$ may have less than $\sigma$ transactions. Also, the rate $\omega$ is only applied when rendering the animation, e.g. if each snapshot $S_i$ yields an image, the stop motion animation will use $\omega/\eta$ images per second to achieve the desired rate of $\omega$ transactions per second. In other words, the transactions in each snapshot are defined by $\sigma$ and $\eta$ and is independent from $\omega$.

@@

\noindent Finally, complementary information may be displayed as a persistent legend that reveals essential information such as $\sigma$, $\eta$, $\omega$ and $T$, and other information that assists the analyst in keeping track of essential features of the current snapshot, e.g. its number of nodes and edges.
Also, node-specific information may be displayed by blinking corresponding cues over the nodes periodically, displaying them with a vertical displacement from the node, or by writing them to a specific widget when requested by the analyst on an interactive interface, i.e. on-demand.
Figure~\ref{fig:final} exemplifies two early Versinus settings.

\begin{figure}[!h]\centering
    \subfloat[]{{\includegraphics[width=.46\textwidth, height=4.3cm]{versinus_} }}%
    \quad
  \subfloat[]{{\includegraphics[width=.46\textwidth, height=4.3cm]{im31} }}%
    \caption{Two frames of dynamic network animations achieved using the Versinus method.
    In (a) the individual degrees in the snapshot are blinking over the nodes, and the nodes are counted in auxiliary lines along the bottom and parallel to the periphery line. Current image/snapshot is 911, with 116 nodes and 262 links.
   In (b), a cleaner version is employed to display a snapshot in which the only hub is the overall hub, the second overall hub is a peripheral node, and most nodes have high clustering coefficient. The legend indicates that the total number of transactions is $T=3,998$, the window size is $\sigma=40$, the step size is $\eta=40$, the transaction rate is $\omega=4$, the current snapshot starts at transaction 1,240, and has 13 nodes and 15 links.
   The low-quality images are typical of the early implementations of Versinus, developed to assist the investigations of the researchers directly related to its development.
   Further information is provided in Section~\ref{sec:des}.
   }%
    \label{fig:final}%
\end{figure}

\subsection{Observation of the evolving network}
Within the settings described, one may notice a number of network characteristics: do nodes (or a specific node) change their sectors often? Is the overall intermediary sector in greater activity with the hubs or periphery?
Does higher clustering occur most often in the less connected nodes?
Are there very unbalanced nodes with respect to in and out degree (or strength)?
Is the number of nodes and edges stable along the network evolution?
In fact, Versinus was partly designed to inspect network attributes due to the remarkable stability of their overall characteristics reported in~\cite{stab}.
In this context, it is natural to ask if such stability is also noticeable in individual nodes or in the connectivity patterns. Although the overall characteristics may be regarded as surprisingly stable, the stability of network components is very network-dependent. E.g., as reported in~\cite{barabasiEvo}, hubs exhibit intermittent activity in social networks, except for the smallest networks analyzed.

\section{Software implementations}\label{sec:imp}
Currently, there are four software implementations of Versinus known to the authors.
For organization and brevity, these are all linked in\footnote{Central repository for Versinus resources such as implementations, public instances of interactive interfaces, and videos: https://github.com/ttm/versinus}. 
Two of them are written in Python: one is a script that renders animations as in the Figure~\ref{fig:final}~(a),
with more cues of local structure. The other is a library that renders animations as in the Figure~\ref{fig:final}~(b), simpler and with more cues of the overall structure, and is capable of rendering sonifications of the dynamic networks, yielding ``audiovisualizations'', i.e. synced sonic and visual mappings of the dynamic networks, as described in Section~\ref{sson}.
The other two implementations are achieved through JavaScript and WebGL.
One is, by default, for static (non-dynamic) networks, implemented as part of the ccNetViz library, that aims at being as computationally inexpensive as possible~\cite{cell}.
The other is an experimental interactive visualization application in which Versinus is combined with other analytical techniques,
such as principal component analysis, overview first - focus on demand, and simultaneous highlighting of corresponding structures across images as required by users.
These last two implementations are illustrated in Figure~\ref{fig:extra}.

\begin{figure}[!h]\centering
    \subfloat[Versinus in ccNetViz]{{\includegraphics[width=.46\textwidth, height=2.5cm]{ccNetViz} }}%
    \qquad
  \subfloat[Experimental interactive visualization]{{\includegraphics[width=.46\textwidth]{visAn} }}%
    \caption{Two implementations of Versinus in JavaScript using WebGL. (a) illustrates the implementation that is available within ccNetViz, a library purposed to being computationally inexpensive. This implementation handles static networks by default. (b) is an image achieved using an experimental interactive visualization application.
    This implementation renders the animations in real time, with multiple controls for inspecting the network structure and controlling the animation, in association with other analysis techniques. The sine oscillates both in the vertical and depth axes, as does the periphery segment.}
    \label{fig:extra}
\end{figure}

\subsection{Network sonification}\label{sson}
The implementation illustrated in Figure~\ref{fig:final} (b) includes a sonification routine for the evolving network.
Results are of limited effectiveness in the sense that the sonification does not enable users to grasp network features with ease,
specially if they are not musically trained.
Nevertheless, it made the (audio)visualization more engaging, as it yields exotic media,
and it is the first time the authors appreciated a sonification of an evolving network (maybe even of any network).
It is possibly a unique multimedia representation of networks.
Accordingly, we understand suitable to provide a short description of the sonification as it was left implemented, i.e. in its final version,
even though a full understanding of the summary requires knowledge of musical theory~\cite{mass}.
Four sonic/musical voices are used, two of them are noises and the other two have defined pitch. The lower noise has central frequency proportional to first hub connectivity,
which is the node positioned at the extreme left on the layout, on the starting point of the sine, and its frequency is updated every two beats.
The higher noise has central frequency proportional to the second hub, the node next to the first hub, and its frequency is updated at each beat.
The third hub connectivity is proportional to the lower note,
which is incident once a beat, and has its note value updated at every beat.
The fourth hub connectivity is proportional to the higher note, which is incident twice every half-beat, is updated at every note, and is only heard on the second half-beat of every tempo.
Figure~\ref{fson} illustrates such disposition of sonic lines (or musical voices).

\begin{figure}[!h]\centering
\includegraphics[width=0.7\textwidth]{sonif}
  \caption{Visual representation of the sonification implemented for the dynamic networks.
  $c_f$ and $f_0$ stands for \emph{center frequency} and \emph{fundamental frequency}, respectively.
  Further information is provided in Section~\ref{sson}.
  }\label{fson}
\end{figure}

\subsection{Animations produced and available}\label{sec:vid}
A number of animations were obtained through the routines described.
They encompass different networks, settings ($\alpha$, $\sigma$, $\eta$,
etc), and different flavors of Versinus.
For organization and brevity, they are available at~\footnote{Central repository for Versinus resources such as implementations, public instances of interactive interfaces, and videos: https://github.com/ttm/versinus}.

\section{A simple case study}\label{visAn}
This section exemplifies the employment of the Versinus network visualization method
within a more elaborate framework through a very simple case study.
Let us consider an interactive visual interface, with which the user may parametrize Versinus and then navigate through the entailed dynamic network.
A most fundamental Versinus parametrization is performed before the visual mapping of the network, and is illustrated in Figure~\ref{env1}.
Initial visualization is illustrated in Figure~\ref{env2}.
Figures~\ref{env3} and~\ref{env4} exemplifies Versinus compliant interactive data visualization procedures: the observation of the nodes in each sector, and characteristics of individual nodes and of snapshots.
The interactive interface is named $ESFNetVis$ (after ``Evolutive Scale Free Network Visualization'').
Directions for accessing the code and operating instances of $ESFNetVis$
are provided in\footnote{Central repository for Versinus resources such as implementations, public instances of interactive interfaces, and videos: https://github.com/ttm/versinus}.

\begin{figure}[!h]\centering
\includegraphics[width=0.7\textwidth]{esfNetVis___}
  \caption{Visual elements for the settings performed before
  the visual mapping of the dynamic network.
  The buttons at the top are for keeping, modifying and accessing settings.
  The widgets in the gray rectangles enables the selection of the network, the first and last messages (transactions) to be considered,
  and the sectorialization method.
  The widgets in the white rectangle are $\sigma$, $\eta$, and $\omega$.
  These settings are compliant with the description available in Section~\ref{sec:des}. The ``Erdös'' setting stands for the achievement of the network sectors by comparing the network against an Erdös-Rényi model, mentioned in Section~\ref{sec:pla}.
  When the red button is clicked, the dynamic network is mapped to the canvas and additional elements, as described in Figure~\ref{env2}.
  }\label{env1}
\end{figure}
